---
title: "RNASeq WorkFlow"
author: "Barry L Hykes Jr"
date: "10/2/2018"
output: html_document
---


###Open a browser and log into RStudio 
```{r Libraries, eval = FALSE}
#load libraries
library("phyloseq"); packageVersion("phyloseq")
library("DESeq2"); packageVersion("DESeq2")
library("ggplot2"); packageVersion("ggplot2")
library("ggrepel"); packageVersion("ggrepel")
library("data.table"); packageVersion("data.table")
```

Read HTSeq output into DESeq2
###The following commands are adopted from the [DESeq2](https://bioconductor.org/packages/release/bioc/manuals/DESeq2/man/DESeq2.pdf) manual

####First we need to read in the files
```{r DESeq, eval=FALSE}
directory <- "/home/genomics/workshop_materials/Transcriptomics/Counts/"
sampleFiles <- list.files(path=directory,patter=".txt")
sampleFiles
```

####Now read in your metadata from the "Conditions.txt" file
```{r, eval = FALSE}
sampleCondition<- read.table("/home/genomics/workshop_materials/Transcriptomics/Conditions.txt",head=TRUE) #file with sample data to be compared
sampleCondition

sampleCondition <- sampleCondition[order(sampleCondition$SampleID),] #We need to order this table by SampleID before merging with the sampleFiles
sampleCondition
```

####combine sample data into a data frame with file names
```{r, eval = FALSE}
sampleTable <- data.frame(sampleName = sampleFiles, fileName = sampleFiles, condition = sampleCondition)
sampleTable #verify that the "fileName" column matches the condition.SampleID column
```

####Now we have our metadata formatted in a way that can be used by DESeq2, but we need to create a data matrix that can be used by the DESeq function
```{r, eval = FALSE}
ddsCounts <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable, directory = directory, design = ~ condition.condition) #creates DESEqDataSet for DESeq function
```

PAUSE 🛑
------------------------------------------------- 

####Before we run the DESeq function, we need to do a final quality control step to make sure there are no batch effects or outliers
```{r, eval = FALSE}
#The data first need to be transformed using the variance stabilizing transformation
vsd <- vst(ddsCounts, blind=FALSE) #variance stabilizing transformation
pcaData <- plotPCA(vsd, intgroup=c("condition.condition"),returnData=TRUE)
percentVar <- round(100 * attr(pcaData,"percentVar"))
head(pcaData)

ggplot(pcaData, aes(PC1, PC2, color=condition.condition))+
  geom_text(mapping = aes(label = name),size=2.5)+
  geom_point(size=2.5) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  xlim(-25,25)+
  coord_fixed()
```

####We can see that the sample "CGCAATTC.txt" performs drastically differently than the rest of the samples, so we need to remove it
```{r, eval = FALSE}
vsd2 <- vsd[,vsd$condition.SampleID != "CGCAATTC.txt"]

pcaData2 <- plotPCA(vsd2, intgroup=c("condition.condition"),returnData=TRUE)
percentVar <- round(100 * attr(pcaData2,"percentVar"))
head(pcaData2)

ggplot(pcaData2, aes(PC1, PC2, color=condition.condition))+
  #geom_text(mapping = aes(label = name),size=1.5)+
  geom_point(size=3) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed()
```

####now we have to remove the outlier(s) from the dds counts matrix since this is what the DESeq function will be run on
```{r, eval = FALSE}
ddsCounts2 <- ddsCounts[,ddsCounts$condition.SampleID != "CGCAATTC.txt"]

#Now we can run the differential expression analysis
dds <- DESeq(ddsCounts2) #DESeq2 command for differential expression test
```

####Extracting DESeq2 Results
Below is a description of the column headers that will appear in the DESeq2 results table:
 * Basemean - mean of normalized counts for all samples
 * log2foldchange
 * lfcse - standard error
 * stat - wald statistic (log2foldchange / lfcse)
 * pvalue 
 * padj - pvalue adjusted for false discovery
####Warning: 
You MUST resist the temptation to use the p-value when evaluating DESeq2 results. The adjusted p-value (padj) accounts for multiple hypothesis testing 
```{r DESeq Results, eval=FALSE}
#text file with ensembl ID's and corresponding gene names
mmGenes = read.table("/home/genomics/workshop_materials/Transcriptomics/MouseGeneTable.txt", header=TRUE, row.names = 1)
mmGenes = as.matrix(mmGenes) #I convert the data frame to a matrix to get a downstream function, cbind, to work properly
head(mmGenes)
```

####extract results table from DESeq2
```{r, eval = FALSE}
res <- results(dds)
res.dds <- res[order(res$padj),] #order table by adj p value
summary(res.dds) #lets take a look at how many genes are up-regulated/down-regulated
  
  # Quick check of factor levels
mcols(res.dds, use.names = TRUE) #this is important for results interpretation
```

We can add the gene names to the results by merging the mmGenes table with the res.dds table
```{r, eval = FALSE}
res.dds = cbind(as(res.dds, "data.frame"), Gene_Name = as(mmGenes[rownames(res.dds), ],"matrix"))
head(res.dds)
```

####Make a new object to add significance boolean for plotting
```{r, eval = FALSE}
resdt.dds = data.table(as(res.dds, "data.frame"),
                 keep.rownames = TRUE)
setnames(resdt.dds, "rn", "Ensembl_ID") 

resdt.dds
resdt.dds[, Significant := padj < .1]
resdt.dds[!is.na(Significant)]
resdt.dds
  
write.table(res.dds,file="/home/genomics/workshop_materials/Transcriptomics/FF_vs_Cre.DESeq.out.txt",quote=F,sep="\t")
```

####make rnk table for Gene Set Enrichment Analysis
```{r, eval = FALSE}
RNK = data.table(Gene_Name = res.dds$Gene_Name, stat = res.dds$stat) #These are 2 columns from our deseq2 output that we need for GSEA
RNK = subset(RNK, stat != "NA")
head(RNK)
write.table(RNK, "/home/genomics/workshop_materials/Transcriptomics/FF_vs_Cre.DESeq.rnk",quote=F,sep="\t", row.names = F)
  
```

Volcano plots are a good way of visualizing the effects of the experimental conditions on the cells
```{r Volcano Plot, eval = FALSE}
volcano = ggplot(
    data = resdt.dds[!is.na(Significant)],
    mapping = aes(x = log2FoldChange,
                  y = -log10(padj),
                  color = Significant,
                  label = Ensembl_ID, label1 = Gene_Name)) +
    theme_bw() +
    geom_point() + 
    geom_point(data = resdt.dds[(Significant)], size = 7, alpha = 0.7) + #Larger circles for the significant values
    theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5)) +
    geom_hline(yintercept = -log10(.1)) +
    ggtitle("DESeq2 Negative Binomial Test Volcano Plot \nProximal Colon Abx vs no Abx") +
    theme(axis.title = element_text(size=12)) +
    theme(axis.text = element_text(size=12)) +
    theme(legend.text = element_text(size=12)) +
    geom_vline(xintercept = 0, lty = 2)
volcano
volcano + xlim(-7,7) #xlim isn't necessary, but I add it here to make the plot symmetrical about the y axix
summary(resdt.dds)

save.image("./Transcriptomics.RData") #it's to to save the image so all the object that have been initiated can be accessed at any time
```

After running GSEA, we may want to make some heat maps. To do that we need to merge HTSeq counts files with Gene list file to make a counts table
```{r, eval = FALSE}
counts = mmGenes #initialize counts table as gene list with gene names and ensemble IDs
#subset sampleFiles to remove the outlier we removed earlier
sampleFiles

#iterate over each HTSeq file
for(i in 1:length(sampleFiles)){ 
  file <- read.table(paste(directory,sampleFiles[i],sep=""),header=FALSE,sep="\t",row.names = 1)

  colnames(file) = c(sub('\\.txt$', '', sampleFiles[i])) #add the file name minus the ".txt" suffix to the header so we can identify each column
  
  dat = subset(file, !(row.names(file)%in%c("__no_feature","__ambiguous","__too_low_aQual","__not_aligned","__alignment_not_unique"))) #remove the extra fields from the counts table
  
 counts = cbind(as(counts,"matrix"),as(dat,"data.frame")) #each time the loop iterates, it will add a new column for the file's counts
  
}
head(counts)
write.table(counts,"/home/genomics/workshop_materials/Transcriptomics/CountsTable.txt",quote=F,sep="\t")
```